---
output: html_document
---
<h3> <center> Appendix </center> </h3>
  
Below are the libraries and data used for this analysis. 

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)

# Loading in the cleaned survey Data
survey_data <- read_csv("/Users/celinekim/Desktop/ps3/survey_data.csv")

# Loading in the cleaned census Data
census_data <- read_csv("/Users/celinekim/Desktop/ps3/census_data.csv")
```

Below is the initial model created for the analysis. The explanatory variables
used in this initial model are is_working, gender, census_region, hispanic,
is_white, is_black, is_american_indian, is_asian_or_pacific_islander, age_group
and us_born. The variables that have been excluded in the final report are
census_region, is_american_indian and is_asian_or_pacific_islander. 

```{r}
# The initial model
initial_model<-glm(as.factor(vote_trump_2020) ~  as.factor(is_working) + 
                     as.factor(gender) + as.factor(census_region) + 
                     as.factor(hispanic) + as.factor(is_white) + 
                     as.factor(is_black) + as.factor(is_american_indian) + 
               as.factor(is_asian_or_pacific_islander) + as.factor(age_group) +
             as.factor(us_born), data=survey_data, family="binomial")
```

Below is the summary data of the initial model that was not included in the 
final report.
```{r}
# Model Results
broom::tidy(initial_model)
```

Below is the logistic regression model used for the analysis in the final 
report.

```{r}
# Creating the Model
model <- glm(as.factor(vote_trump_2020) ~  as.factor(is_working) + 
             as.factor(gender) + as.factor(hispanic) + as.factor(is_white) + 
             as.factor(is_black) + as.factor(age_group) + as.factor(us_born), 
           data=survey_data, family="binomial")
```

Below is the post-stratification analysis.
```{r}
# Here I will perform the post-stratification calculation
census_data$estimate <-
  model %>%
  predict(newdata = census_data)

census_data <- census_data %>%
  mutate(trump_predict_prop = estimate*cell_prop_of_division_total)

logit2prob <- function(logit){
  odds <- exp(logit)
  prob <- odds / (1 + odds)
  return(prob)
}

trump_predict_overall <- sum(census_data[, 'trump_predict_prop'], na.rm=TRUE)
trump_predict_overall_prob <- logit2prob(trump_predict_overall)
```


#### Table 3: Results of Additional Analysis Performed Per State Basis
```{r, echo=FALSE, message=FALSE}
additional_data <-
census_data %>%
  mutate(trump_predict_prop_state = 
           estimate*cell_prop_of_division_total_state) %>%
  group_by(state) %>%
  summarise(trump_predict_state = sum(trump_predict_prop_state), 
            trump_predict_state_prob = logit2prob(trump_predict_state))

print(additional_data, n = 50)
```

