---
title: "Predicting the Proportion of Voters in Favour of Donald Trump in the 2020 American Federal Election"
author: "Celine Kim (Student Number: 1001251147)"
date: "2 November, 2020"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
library(tidyverse)
library(broom)

# Loading in the cleaned survey Data
survey_data <- read_csv("./outputs/survey_data.csv")

# Loading in the cleaned census Data
census_data <- read_csv("./outputs/census_data.csv")
```

# Model
In order to predict the overall popular vote of the 2020 American federal
election, I will use a logistic regression model with post-stratification. I am
interested in predicting the proportion of voters who will vote for Donald Trump
so I will be modelling a response variable on whether an individual will vote
for Trump or not. The model specifics and calculations used in the
post-stratification technique will be explained in the following subsections of
this report.

## Model Specifics
Since whether one will vote for Donald Trump or not in the 2020 American federal
election is a binary outcome, a logistic regression model will be used to model
this binary response variable. The aspects focused in this model are the
employment status, gender, census region, racial ethnicity, age and birthplace
of the voter. Firstly, according to the report "An examination of the 2016
electorate, based on validated voters" conducted by the Pew Research Center of
U.S. Politics and Policy, "88% of Trump's supporters were Whites, and the age
group of Trump's voters were higher than Clinton's" which shows that age and
racial ethnicity or birthplace influence voting decisions. In addition, a more
recent 2020 report by the same source states that "attitudes of voters  on race
and gender in 2020 are more divided than in 2016". Also, from a 2016 article by
the Washington Post, there is a proposition that "Trump supporters are from
communities with high unemployment rates" which shows that employment status and
region of voters are another crucial factors to consider when predicting the
result of the presidential election.

Thus, in this model, the following variables are used. 

Let $x_{emp}$ represent the employment status where 1 represents that the voter 
is employed and 0 otherwise. 

Let $x_{male}$ be the gender of the voter. This is a binary categorical variable
where 1 represents that the voter is male and 0 if female. 

Let $x_{his}$ be a binary categorical variable where 1 represents that the
voter is of Hispanic, Latino or Spanish origin and 0 otherwise. 

Let $x_{white}$ be a binary categorical variable where 1 represents that the
voter's racial ethnicity is White. Let $x_{black}$ be a binary categorical 
variable where 1 represents that the voter's racial ethnicity is Black or 
African American.

Let $x_{30to44}$, $x_{45to59}$, $x_{60plus}$ be binary categorical variables 
where 1 represents that the voter belongs to the age group between 30 to 44, 
45 to 59, 60 plus, respectively.

Let $x_{bpl}$ be a binary variable that represents the birthplace of the voter 
where 1 represents that the voter was born in the United States of America.

The event of interest in this model is whether the voter will vote for for 
Donald Trump in the 2020 American federal election. Then let $p$ be the 
probability that one will vote for Donald Trump in this election. In this model,
$\log{\frac{p}{1-p}}$ is the log odds of one voting for Trump, the response 
variable.

Then this model in mathematical notation can be written as: 

$$\log{\frac{p}{1-p}} = \beta_0 + \beta_1x_{emp} + \beta_2x_{male} + \beta_3x_{his} + \beta_4x_{white} + \beta_5x_{black} + \beta_6x_{30to44} + \beta_7x_{45to59} + \beta_8x_{60plus} + \beta_9x_{bpl}$$

$\beta_0$ is the intercept which represents when $x_{emp}$, $x_{male}$, 
$x_{his}$, $x_{white}$, $x_{black}$, $x_{30to44}$, $x_{45to59}$, $x_{60plus}$ 
and $x_{bpl}$ are all 0. In this specific model, this situation corresponds to 
the log odds of one voting for Trump when the voter is a foreign-born, 
unemployed female who is not of Hispanic, White or African American origin and 
in the age group between 18 to 29 (since voters of age of 17 and under has been 
excluded).

$\beta_1$ to $\beta_9$ are coefficients that represent the change in log odds
for every one unit increase in the respective explanatory variables. Since all
explanatory variables are binary variables, one unit increase in this case
represents presence of the corresponding condition.

In the initial statistical model that can be found in the Appendix, variables
such as the census region of the voter and whether the voter's racial ethnicity
is American Indian, Asian or Pacific Islander were also included. However, using
the p-value base of 0.05, I decided to remove such variables from the model.
This is a sensible decision because people of Hispanic or African American
origin in the U.S. are those in conflict with Trump while little or no conflict
exists between Trump and Asians, Native Americans or Pacific Islanders.


```{r, echo=FALSE}
# Creating the Model
model <- glm(as.factor(vote_trump_2020) ~  as.factor(is_working) + 
             as.factor(gender) + as.factor(hispanic) + as.factor(is_white) + 
             as.factor(is_black) + as.factor(age_group) + as.factor(us_born), 
           data=survey_data, family="binomial")
```


## Post-Stratification 

Post-stratification is a statistical analysis technique that involves dividing
data into small demographic cells, adding those cell-level estimates and
extrapolating to create a population-level estimate in order to predict the
response variable. This technique is useful because it enables correction for
non-probability sampling errors and accurate estimate for population level
predictions. Therefore, I performed a post-stratification analysis to estimate
the "nationwide" proportion of voters who will vote for Donald Trump.

I created bins based on state, is_working, gender, hispanic, is_white, is_black,
age_group and us_born because these were the parameters (that were also
available in the census data) used to develop the logistic regression model. The
variable "state" is included to perform further analysis where we make
predictions on the proportion of voters per state on who will vote for Trump
(please check the Additional Analysis Per State Basis). Initially, I wanted to
include the voting data from the 2016 presidential election, preference of news
source such as CNN or Fox, primary party that one follows as these data were
available in the Democracy Fund and UCLA Nationscape ‘Full Data Set’, but the
census data from the American Community Surveys (ACS) did not include such
information so they were removed. These factors were originally considered
because according to an article called "Fox News was No. 1 news source for Trump
voters", "40% of those who voted for Trump relies on Fox News as the main source
of news in the 2016 election" which shows that news source is another aspect to
consider.


```{r, echo=FALSE}
# Here I will perform the post-stratification calculation
census_data$estimate <-
  model %>%
  predict(newdata = census_data)

census_data <- census_data %>%
  mutate(trump_predict_prop = estimate*cell_prop_of_division_total)

logit2prob <- function(logit){
  odds <- exp(logit)
  prob <- odds / (1 + odds)
  return(prob)
}

trump_predict_overall <- sum(census_data[, 'trump_predict_prop'], na.rm=TRUE)
trump_predict_overall_prob <- logit2prob(trump_predict_overall)
```

#### Additional Analysis Per State Basis
The United States presidential election is an indirect voting system where the
citizens vote for the members of the Electoral College instead of directly
choosing between the candidates. After the general election where citizens cast
a vote, the electors then vote based on the state results. Finally, the
presidential candidate is decided based on the results of votes of the electoral
college. Thus, a more accurate prediction would require calculations per state
basis. Therefore, I grouped by state and included it as a bin to conduct
post-stratification analysis on vote per state.

The result of this additional analysis is in Table 3 of the Appendix.

```{r, include=FALSE, warning=FALSE, message=FALSE}
census_data %>%
  mutate(trump_predict_prop_state = 
           estimate*cell_prop_of_division_total_state) %>%
  group_by(state) %>%
  summarise(trump_predict_state = sum(trump_predict_prop_state), 
            trump_predict_state_prob = logit2prob(trump_predict_state))
```

# Results

#### Table 1: Summary of Logistic Regression Model

Below is the summary of the logistic regression model used in this report.
```{r, echo=FALSE}
broom::tidy(model)
```
Since the p-value of variables in the logistic regression model are all below 
0.05, this provides grounds to reject the null hypotheses that $\beta_0$
to $\beta_9$ = 0.

This model can be mathematically expressed as:

$$\log{\frac{p}{1-p}} = -1.9063 + 0.2089x_{emp} + 0.3745x_{male} -0.3208x_{his} 
+ 0.5188x_{white} - 1.4687x_{black} + 0.5364x_{30to44} + 0.6814x_{45to59} $$ \n
$+ 0.7100x_{60plus} + 0.4077x_{bpl}$

#### Table 2: Results from Post-Stratification Analysis
```{r, echo=FALSE}
trump_predict_overall
trump_predict_overall_prob
```

The results from the post-stratification analysis are that 
$\log{\frac{p}{1-p}}$ = -0.4973671 and $\hat{y}^{PS}$ = 0.3781596.

Using the post-stratification analysis of the proportion of voters in favour of
Trump modelled by a logistic regression model, I estimate that the proportion of
voters in favour of voting for Trump to be 0.3782. The logistic regression model
used in this analysis accounts for the employment status, gender, racial
ethnicity, and age group of the voter and whether the voter was born in the U.S.
or not.

The result of the prediction was trump_predict_overall which is the log odds of
voting for Donald Trump. Then, I converted the log odds to the probability by
calculating exp(trump_predict_overall) /(1+exp(trump_predict_overall)).

Table 3 of the Appendix is the result from the additional analysis performed to
estimate the proportion of voters in favour of Trump per state basis. As shown
in Table 3 of the Appendix, the log odds of voting for Donald Trump for all
states is less than 0, meaning that the proportion will be less than 0.5 in all
states.

#### Figure 1: Barplot of The Proportion of Voters in Favour of Trump per State Basis

Below is the barplot of the proportion of voters in favour of Trump per state 
basis using the result of the additional analysis.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
additional_data <-
  census_data %>%
  mutate(trump_predict_prop_state = 
           estimate*cell_prop_of_division_total_state) %>%
  group_by(state) %>%
  summarise(trump_predict_state = sum(trump_predict_prop_state), 
            trump_predict_state_prob = logit2prob(trump_predict_state))

barplot(trump_predict_state_prob~state, data=additional_data, ylim=c(0,1),
        las=2,cex.names=0.5,
        main="The Proportion of Voters in Favour of Trump per State Basis")
```

# Discussion

From the survey data obtained from the Democracy Fund and UCLA Nationscape ‘Full
Data Set’, I selected the employment status, gender, census region, racial
ethnicity, age and birthplace as factors since they were shown to have influence
on voting decisions. Then, I built a logistic regression model that predicts
voting in favour of Donald Trump using these data at an individual level. Then,
we conducted post-stratification analysis using census data from 2018, to
predict proportion of voters nationwide and per state basis (used in the
Additional Analysis Per State Basis) in favour of Donald Trump.

In conclusion, Donald Trump will lose not only the overall nationwide popular
vote of the 2020 American federal election but also in every state. As shown in
the nationwide election results of Table 2, only 0.379 will vote for Trump, and
in calculations based on each state in Table 3 of the Appendix, the proportion
in each state is less than 0.5. The maximum proportion is 0.4574 in Maine. The
proportion was between 0.4 and 0.5 in about 50% of the states, and the other 50%
of the states had proportions below 0.4. Thus, it is not likely that Trump will
win. Therefore, regardless of which voting system the U.S. is using, it is
likely that Donald Trump will lose the 2020 presidential election.

## Weaknesses
There are other important parameters that have great influence on predicting the
result of the 2020 presidential election including the voting data from the 2016
election, the main preference for news sources as mentioned before, opinions on
how the current government is coping with the coronavirus outbreak and personal
thoughts on policies such as the Green New Deal and deportation of unregistered
immigrants. The citizens' opinions on such regulations have great influence on
the voting decisions. For example, while the Green New Deal is supported by the
Democratic Party, Trump opposes the regulation. According to an article by the
Washington Post (cited in #7 of the Reference Section), the opinion is divided
on whether the country needs to cut down on the use of fossil fuels between
Democrats and Republicans and "most Americans oppose paying trillions of dollars
by tax to reach the goals of the regulation". However, these variables were not
included in the analysis because such data were not available in census data
provided by ACL. Even though these variables might have greater impact than the
variables we studied, since the census data includes more objective information
that does not inquire specifically on voting or political issues, I was not able
to utilize these parameters. Especially, the vote rate per state in the previous
election might be a very important factor when calculating the proportion of
voters per state. In addition, since this model uses data based on June 25th,
2020, the model does not take into account recent changes that happened after
June 25th including Trump getting COVID-19 and election debates between
September and October which might have influenced or changed people's voting
decisions.

## Next Steps
If the data on the voting rates per age group in each state is available, this
information can be used to re-adjust the proportion per bin because the
population proportion does not represent the proportion of those who actually
vote. This additional information can yield a more accurate prediction. In
addition, since the actual data will be available after the election occurs,
post hoc analysis can be performed to examine if each variable used in this
analysis are critical variables that influence the voting trend and analyze to
improve the model for better prediction in the future elections.


# References
1. Tausanovitch, C., & Vavreck, L. (2020). Democracy Fund + UCLA 
  Nationscape, October 10-17, 2019 (version 20200131). Retrieved from             
  https://www.voterstudygroup.org/downloads?key=b5385d6e-a729-4a43-8ab8-4c6d6fca3100. 

2. Steven Ruggles, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer, Jose 
  Pacas and Matthew Sobek. (2020) IPUMS USA: Version 10.0 [dataset]. 
  Minneapolis, MN: IPUMS, 2020. https://doi.org/10.18128/D010.V10.0

3. Pew Research Center. (2018, August 09). An examination of the 2016 
  electorate, based on validated voters. Retrieved October 29, 2020, from 
  https://www.pewresearch.org/politics/2018/08/09/an-examination-of-the-2016-electorate-based-on-validated-voters/

4. Pew Research Center. (2020, September 10). Voters' Attitudes About Race and 
  Gender Are Even More Divided Than in 2016. Retrieved October 30, 2020, from         
  https://www.pewresearch.org/politics/2020/09/10/voters-attitudes-about-race-and-gender-are-even-more-divided-than-in-2016/

5. Aytaç, S., & Rau, E. (2019, April 18). Trump supporters vastly overestimate 
  unemployment - and they blame politicians for it. Retrieved October 29, 2020, 
  from 
  https://www.washingtonpost.com/news/monkey-cage/wp/2016/11/02/trump-supporters-vastly-overestimate-unemployment-and-they-blame-politicians-for-it/ 

6. Sutton, K. (2017, January 18). Fox News was No. 1 news source - for Trump 
  voters. Retrieved November 01, 2020, from     
  https://www.politico.com/blogs/on-media/2017/01/study-fox-news-is-no-1-news-source-for-trump-voters-233773

7. Clement, S., & Grandoni, D. (2019, December 04). Americans like Green New 
  Deal's goals, but they reject paying trillions to reach them. Retrieved 
  November 01, 2020, from 
  https://www.washingtonpost.com/climate-environment/2019/11/27/americans-like-green-new-deals-goals-they-reject-paying-trillions-reach-them/

8. Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source 
  Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

9. David Robinson, Alex Hayes and Simon Couch (2020). broom: Convert Statistical 
  Objects into Tidy Tibbles. R package version 0.7.0. 
  https://CRAN.R-project.org/package=broom
  
  


